The imprint of global magma oceans on exoplanet demographics

### 2023-04-11 src/tex/ms.tex
review introduction

An example from the Solar System for the potential significance of these early stages are the divergent atmospheric evolutions of Venus and Earth~\citep{2019JGRE..124.2015K,2021JGRE..12606643K}.
While having accreted from a similar mass reservoir~\citep{2020plas.book..287R,2020SSRv..216...55K,2020SSRv..216...27M,2020plas.book....3Z} and despite their similar bulk properties, they evolved into planets with very different environmental conditions on their surfaces~\citep{1982Sci...216..630D,Kasting1988,Hamano2013,Kane2014,2020JGRE..12506276W,Turbet2021}.
At the formation of the Solar System, both planets underwent a giant impact phase~\citep{2020plas.book..287R,2020NatGe..13..265G,Liu2022} that melted their mantles, leading to a magma ocean stage~\citep{2012AREPS..40..113E,2018RSPTA.37680109S,2022arXiv220310023L}.
Magma ocean states play a substantial role in establishing the geophysical and climatic regimes of rocky planets \citep{2020ChEG...80l5594F}, because atmospheric volatiles are highly soluble in magmatic liquids~\citep{2021ApJ...909L..22K,Dorn2021}.
Due to these similar early formation phases, it was commonly assumed that the divergence of Venus and Earth -- in particular Venus's water loss -- occurred late in their evolution~\citep{Elkins-Tanton2013}.

However, no late desiccation needs to be invoked if a strongly infrared-absorbing, condensable species such as water is present on the planet \citep{Hamano2013}, which is supported by isotopic and petrological evidence from meteoritic samples \citep{2018SSRv..214...36A,2022Natur.611..245B}.
Once the incident stellar flux exceeds a threshold value, a non-linear transition from mixed condensed and volatile to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,2013NatGe...6..661G,2015ExA....40..449L,2017JGRE..122.1458S}.
This runaway greenhouse state prevents the planet's rapid cooling and can extend the magma ocean stage to hundreds of \SI{}{\mega\year} \citep{2016ApJ...829...63S,2021AsBio..21.1325B}, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen \citep{2015AsBio..15..119L,2013ApJ...778..154W,2014ApJ...785L..20W}.
%Because hydrodynamic escape driven by stellar XUV decreases with time, the solidification timescale can become comparable to the main-sequence lifetime of the star with a sufficient water inventory~\citep{2013Natur.497..607H,2015ApJ...806..216H,2022arXiv220310023L}.
Venus' atmospheric composition is consistent with the planet having gone through this phase \citep{2020NatGe..13..265G}, although the past presence or absence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016,2019JGRE..124.2015K,2021JGRE..12606643K,Turbet2021,2023PNAS..12009751W}.

@@

Of particular relevance is that the radiation-induced transition between a runaway greenhouse state and a temperate climate is thought to occur at a relatively sharp threshold instellation~\citep{2010ppc..book.....P,2013ApJ...765..131K}. Consequently, the instellation at which the runaway greenhouse transition occurs is aptly considered to be the inner boundary of the habitable zone, whose prevalent definition refers to the possibility of liquid water on an Earth-like planet with a oxidized CO$_2$/H$_2$O/N$_2$-rich atmospheres~\citep[][]{Kasting1993,Kopparapu2013,Kopparapu2014}, but can be influenced for different internal redox states and volatile compositions \citep{2011ApJ...734L..13P,2017ApJ...837L...4R,2018ApJ...858...72R,2019ApJ...875...31K,2020ApJ...896..115G,2022JGRE..12707456G,2023ApJ...942L..20H}.
The fundamental concept of a habitable zone dates back centuries~\citep{Newton1687,Whewell1858,Shapley1953,Huang1959}, and its modern form has proven popular in the planetary literature\footnote{At the time of writing, a search of the term ``habitable zone'' in the titles and abstracts of refereed articles in NASA's Astrophysics Data System returned $\sim\num{1700}$ results.}. However, it should be emphasized that the habitable zone, as it stands today, is solely a concept based on theoretical predictions, and its validity continues to be debated~\citep[e.g.,][]{Cockell2016,Moore2017}.
The question naturally emerges if a planetary habitable zone -- in its common form with boundaries simply defined by the received stellar irradiation -- is a predicitive theoretical concept and how the habitable zone hypothesis can be tested observationally.
Observational tests being considered include searches for evidence of liquid water conveyed by ocean glint~\citep{Robinson2010,Lustig-Yaeger2018} or water vapor in planetary atmospheres~\citep{Suissa2020}.
A different approach relies on comparative planetology: By aiming for a statistical detection in a planet population, degeneracies and ambiguousities in making statements about the habitability of individual planets may be overcome~\citep{Checlair2019}.
Tests suggested in the literature include determining, for a range of orbital distances, atmospheric H$_{2}$O and CO$_2$ abundances, planetary albedos ~\citep{Bean2017,Bixel2021}, or colors~\citep{Crow2011}, or testing the relationship between CO$_2$ partial pressure and incident flux~\citep{Lehmer2020}.
All these tests require surveying a large enough sample of terrestrial-sized planets with next-generation instruments, rendering them out of reach in the immediate future.

@@

\section{Runaway Greenhouse Model}\label{sec:met_rghmodel}
%\todo[inline]{TIM: please review} DONE
%\todo[inline]{
%\textbf{Main assumptions:}\\
%- every planet (modulo $f_\mathrm{rgh}$) has \textit{some} water. Enough for runaway GH under the right conditions.  \\
%- Limitation to small planets without significant atmospheric envelopes.\\
%- we ignore water loss by H$_2$O photolysis in the upper atmosphere and subsequent H escape, which would eventually cool even planets within the runaway greenhouse regime~\citep{Lichtenberg2022}.\\
%- we ignore tidal heating, which could extend the magma ocean phase of close-in planets (and change their orbits via tidal orbital evolution)( Jackson et al., 2008; Barnes et al., 2013 (?)).\\
%}
The climate state of a planet has a direct influence on its apparent size measured by transit photometry~\citep{Turbet2019,Turbet2020,Mousis2020,2021ApJ...914...84A}. With even a fraction of the Earth's water inventory, a super-runaway planet will be in a global magma ocean state \citep{2021JGRE..12606711L,2021ApJ...919..130B}. We use predictions on transit atmospheric thickness from geophysical models to derive the change in transit radius and bulk density that planets with instellation-induced runaway greenhouse climates experience, depending on the thermal stratification of the atmosphere and dissolution of water into the underlying magma ocean \citep{Dorn2021,2023Icar..39015265S}.

@@

\subsubsection{Simplifications in the atmosphere model}
%\todo[inline]{TIM: your expertise will be valuable here.\\
%Mention we only consider instellation-induced rgh.\\
%- we ignore water loss by H$_2$O photolysis in the upper atmosphere and subsequent H escape, which would eventually cool even planets %within the runaway greenhouse regime~\citep{Lichtenberg2022}.\\
%- we ignore tidal heating, which could extend the magma ocean phase of close-in planets (and change their orbits via tidal orbital evolution)( Jackson et al., 2008; Barnes et al., 2013 (?)).\\
%- assumption of a single albedo. ``if there is a range of albedos, will that wash away the signal?''
%}
% >>>>>>>>>>
% TL: I just wrote a review article about this exact problem (magma ocean atmosphere exchange) with Yamila Miguel, hopefully soon on arXiv. Honestly, I think we keep it short here, otherwise the paper will get way too long and diluted. I referenced the latest reviews on this now, and we can cite Lichtenberg & Miguel 23 as soon as it's on arXiv for a much much longer discussion.
% >>>>>>>>>>
As illustrated in Fig.~\ref{fig:radiusevolution}, evolutionary processes and a background population of planets unable to sustain a steam atmosphere complicate the classification of a planet into the super- or sub-runaway category. For instance, both a Hydrogen/Helium-dominated ``sub-Neptune'' and a terrestrial planet currently hosting a steam atmosphere will appear as a planet in our runaway greenhouse category, diluting the expected demographic signature. Further, a planet in a runaway greenhouse state will eventually lose its steam atmosphere and move to the sub-runaway greenhouse category on a timescale of $\SI{e8}{\year}$ for a planet with one Earth ocean of water~\citep[][]{Watson1981}. Developing more sophisticated models of the coupled interaction between atmosphere and interior for an arbitrary planetary composition remains a long-term goal of the community \citep{2021SSRv..217...22G,2022arXiv220310023L,2022ARA&A..60..159W}.
We marginalized over these degeneracies by introducing the dilution factor $f_\mathrm{rgh}$, but in a real survey contextual information about a planet's age or likely formation history could be taken into account.

>>>>

An example from the Solar System for the potential significance of these early stages are the divergent atmospheric evolutions of Venus and Earth~\citep{2019JGRE..124.2015K,2021JGRE..12606643K}.
While having accreted from a similar mass reservoir~\citep{2020plas.book..287R,2020SSRv..216...55K,2020SSRv..216...27M,2020plas.book....3Z} and despite their similar bulk properties, they evolved into planets with very different environmental conditions on their surfaces~\citep{1982Sci...216..630D,Kasting1988,Hamano2013,Kane2014,2020JGRE..12506276W,Turbet2021}.
At the formation of the Solar System, both planets underwent a giant impact phase~\citep{2020plas.book..287R,2020NatGe..13..265G,Liu2022} that melted their mantles, leading to a magma ocean stage~\citep{2012AREPS..40..113E,2018RSPTA.37680109S,2022arXiv220310023L}.
Magma ocean states play a substantial role in establishing the geophysical and climatic regimes of rocky planets~\citep{2020ChEG...80l5594F} because atmospheric volatiles are highly soluble in magmatic liquids~\citep{2021ApJ...909L..22K,Dorn2021}.
Due to these similar early formation phases, it was commonly assumed that the divergence of Venus and Earth -- in particular Venus's water loss -- occurred late in their evolution~\citep{Elkins-Tanton2013}.

However, no late desiccation needs to be invoked if a strongly infrared-absorbing, condensable species such as water is present on the planet~\citep{Hamano2013}, which is supported by isotopic and petrological evidence from meteoritic samples \citep{2018SSRv..214...36A,2022Natur.611..245B}.
Once the incident stellar flux exceeds a threshold value, a non-linear transition from mixed condensed and volatile to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,2013NatGe...6..661G,2015ExA....40..449L,2017JGRE..122.1458S}.
This runaway greenhouse state prevents the planet's rapid cooling and can extend the magma ocean stage to hundreds of \SI{}{\mega\year} \citep{2016ApJ...829...63S,2021AsBio..21.1325B}, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen \citep{2013ApJ...778..154W,2014ApJ...785L..20W,2015AsBio..15..119L}.
%Because hydrodynamic escape driven by stellar XUV decreases with time, the solidification timescale can become comparable to the main-sequence lifetime of the star with a sufficient water inventory~\citep{2013Natur.497..607H,2015ApJ...806..216H,2022arXiv220310023L}.
Venus' atmospheric composition is consistent with the planet having gone through this phase \citep{2020NatGe..13..265G}, although the past presence or absence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016,2019JGRE..124.2015K,2021JGRE..12606643K,Turbet2021,2023PNAS..12009751W}.

@@

Of particular relevance is that the radiation-induced transition between a runaway greenhouse state and a temperate climate is thought to occur at a relatively sharp threshold instellation~\citep{2010ppc..book.....P,2013ApJ...765..131K}.
Consequently, the instellation at which the runaway greenhouse transition occurs is aptly considered to be the inner boundary of the habitable zone.
Its prevalent definition refers to the possibility of liquid water on an Earth-like planet with an oxidized CO$_2$/H$_2$O/N$_2$-rich atmospheres~\citep[][]{Kasting1993,Kopparapu2013,Kopparapu2014}, but it can be influenced by different internal redox states and volatile compositions~\citep{2011ApJ...734L..13P,2017ApJ...837L...4R,2018ApJ...858...72R,2019ApJ...875...31K,2020ApJ...896..115G,2022JGRE..12707456G,2023ApJ...942L..20H}.
The fundamental concept of a habitable zone dates back centuries~\citep{Newton1687,Whewell1858,Shapley1953,Huang1959}, and its modern form has proven popular in the planetary literature\footnote{At the time of writing, a search of the term ``habitable zone'' in the titles and abstracts of refereed articles in NASA's Astrophysics Data System returned $\sim\num{1700}$ results.}.
However, it should be emphasized that the habitable zone, as it stands today, is solely a concept based on theoretical predictions, and its validity continues to be debated~\citep[e.g.,][]{Cockell2016,Moore2017,Tuchow2023}.
The question naturally emerges if a planetary habitable zone -- in its common form with boundaries simply defined by the received stellar irradiation -- is a predictive theoretical concept and how the habitable zone hypothesis can be tested observationally.

Observational tests being considered include searches for evidence of liquid water conveyed by ocean glint~\citep{Robinson2010,Lustig-Yaeger2018} or water vapor in planetary atmospheres~\citep{Suissa2020}.
A different approach relies on comparative planetology: By aiming for a statistical detection in a planet population, degeneracies and ambiguousities in making statements about the habitability of individual planets may be overcome~\citep{Checlair2019}.
Tests suggested in the literature include determining, for a range of orbital distances, atmospheric H$_{2}$O and CO$_2$ abundances, planetary albedos ~\citep{Bean2017,Bixel2021}, or colors~\citep{Crow2011}, or testing the relationship between CO$_2$ partial pressure and incident flux~\citep{Lehmer2020}.
All these tests require surveying a large enough sample of terrestrial-sized planets with next-generation instruments, rendering them out of reach in the immediate future.

@@

\section{Runaway Greenhouse Model}\label{sec:met_rghmodel}
%\todo[inline]{TIM: please review} DONE
%\todo[inline]{
%\textbf{Main assumptions:}\\
%- every planet (modulo $f_\mathrm{rgh}$) has \textit{some} water. Enough for runaway GH under the right conditions.  \\
%- Limitation to small planets without significant atmospheric envelopes.\\
%- we ignore water loss by H$_2$O photolysis in the upper atmosphere and subsequent H escape, which would eventually cool even planets within the runaway greenhouse regime~\citep{2022arXiv220310023L}.\\
%- we ignore tidal heating, which could extend the magma ocean phase of close-in planets (and change their orbits via tidal orbital evolution)( Jackson et al., 2008; Barnes et al., 2013 (?)).\\
%}
The climate state of a planet has a direct influence on its apparent size measured by transit photometry~\citep{Turbet2019,Turbet2020,Mousis2020,2021ApJ...914...84A}. With even a fraction of the Earth's water inventory, a super-runaway planet will be in a global magma ocean state \citep{2021JGRE..12606711L,2021ApJ...919..130B}. We use predictions on transit atmospheric thickness from geophysical models to derive the change in transit radius and bulk density that planets with instellation-induced runaway greenhouse climates experience, depending on the thermal stratification of the atmosphere and dissolution of water into the underlying magma ocean \citep{Dorn2021,2023Icar..39015265S}.

@@

\subsubsection{Simplifications in the atmosphere model}
%\todo[inline]{TIM: your expertise will be valuable here.\\
%Mention we only consider instellation-induced rgh.\\
%- we ignore water loss by H$_2$O photolysis in the upper atmosphere and subsequent H escape, which would eventually cool even planets %within the runaway greenhouse regime~\citep{2022arXiv220310023L}.\\
%- we ignore tidal heating, which could extend the magma ocean phase of close-in planets (and change their orbits via tidal orbital evolution)( Jackson et al., 2008; Barnes et al., 2013 (?)).\\
%- assumption of a single albedo. ``if there is a range of albedos, will that wash away the signal?''
%}
% >>>>>>>>>>
% TL: I just wrote a review article about this exact problem (magma ocean atmosphere exchange) with Yamila Miguel, hopefully soon on arXiv. Honestly, I think we keep it short here, otherwise the paper will get way too long and diluted. I referenced the latest reviews on this now, and we can cite Lichtenberg & Miguel 23 as soon as it's on arXiv for a much much longer discussion.
% >>>>>>>>>>
As illustrated in Fig.~\ref{fig:radiusevolution}, evolutionary processes and a background population of planets unable to sustain a steam atmosphere complicate the classification of a planet into the super- or sub-runaway category. For instance, both a Hydrogen/Helium-dominated ``sub-Neptune'' and a terrestrial planet currently hosting a steam atmosphere will appear as a planet in our runaway greenhouse category, diluting the expected demographic signature. Further, a planet in a runaway greenhouse state will eventually lose its steam atmosphere and move to the sub-runaway greenhouse category on a timescale of $\SI{e8}{\year}$ for a planet with one Earth ocean of water~\citep[][]{Watson1981}. Developing more sophisticated models of the coupled interaction between atmosphere and interior for an arbitrary planetary composition remains a long-term goal of the community \citep{2021SSRv..217...22G,2022arXiv220310023L,2022ARA&A..60..159W}.
We marginalized over these degeneracies by introducing the dilution factor $f_\mathrm{rgh}$, but in a real survey contextual information about a planet's age or likely formation history could be taken into account.
